Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-12-26 demystify.tex
Updates from Overleaf
updates

Note that most places only have deadlines for the applicant, but are very flexible with the letter writers.  In many cases your LOR writers \emph{are not given any deadline}. \autoref{fig:fig:lor-invitations} lists several LOR invitation emails I received from various universities in the past few years.

@@

    .. is applying for admission into the Computer Science (Ph.D.) program in the Graduate School at the University of Massachusetts Amherst and has listed you as a reference.

@@

    \textbf{Sample 2:}

    Dear ..

@@

    \textbf{Sample 3:}

    .. is applying to the University of Nebraska-Lincoln and has named you as a recommender. \textbf{This applicant has waived the right to view your recommendation.}

@@

    \textbf{Sample 2:}

    Dear Recommender,

@@

Many international students mistakenly believe that their school is well-known, but in fact, it is not, e.g., many Vietnamese students believe that ``Bach Khoa'' is well-known, but it is not (and their confusing acronyms HUST and HCMUT only make it worse).  So if you think your school is well-known, mention it (and back up with some evidence like rankings or awards) in your SOP.  You can ask your LOR writers to talk about your school as well (\autoref{sec:help-your-LOR-writers}).
Of course, if you're interested in working with Vietnamese, consider  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{CS programs in the US that have Vietnamese professors}. %It might also be helpful to have your CS dept to put itself on CSRankings so that others know about the dept, its faculty, and research.

@@

\paragraph{Locally} Start looking for research opportunities at your institution.
If you did well or liked a class, you can check with the professor of that class for research opportunities.
You can also go to the department directory and then professors' websites and see if they are looking for undergraduate researchers.
Even if they say they are looking for graduate students, you can still contact them and ask.

There might be also special programs for undergraduate research from the university, e.g., Mason has the OSCAR program, UNL has UCARE, and NSF has REUs (for US citizens and permanent residents).

@@

\paragraph{Online Research Communities and Open Source Contributions} Places such as GitHub offer many research opportunities, even though they might not appear so (e.g., they do not explicitly mention research).
In many cases, professors or research labs put their projects on GitHub. For example, my research group, \href{https://github.com/dynaroars/}{Dynaroars}, has many open-source projects that undergrads can contribute to.

@@

\paragraph{Virtual Research Opportunities} This is less common but several places offer virtual internships and research programs aimed at providing hands-on research experience. These programs often involve working remotely under the guidance of experienced mentors and collaborating with a team of fellow researchers. For instance, \href{https://docs.google.com/forms/d/1btIwt4HwjyKMOUk-EMy3rbkfWzFxv2lNrMm_zkd0pA4/viewform?edit_requested=true}{UIUC+ Summer Undergraduate Research in Software Engineering}  offers an unpaid remote internship for software engineering students all over the world to collaborate with mentors from University of Illinois at Urbana Champaign.

@@

\begin{table}[h]
  \centering
  \small
  \caption{The top 50 CS programs in the US from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system  based on faculty publications at top CS conferences. (\href{https://csrankings.org}{CSRankings}, Jan. 2024). \red{$^*$} indicates that the university has \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese prof.} that can advise CS PhD students.}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon & 26 & Duke University \\
    2 & Univ. of Illinois at Urbana-Champaign\red{$^*$}  & 27 & Univ. of California - Santa Barbara \\
    3 & Univ. of California-San Diego & 28 & Rutgers University\red{$^*$} \\
    4 & Georgia Institute of Technology & 29 & Univ. of California - Riverside\\
    5 & MIT                            & 30 & Pennsylvania State University  \\
    6 & Univ. of California - Berkeley& 31 & Northwestern University\\
    7 & University of Michigan - Ann Arbor\red{$^*$}   & 32& George Mason University\red{$^*$}\\
    8 & University of Washington      &33 &  Texas A\&M University\red{$^*$} \\
    9 &  Stanford University  &34&  University of Utah \\
    10 & Cornell University  & 35 &  North Carolina State University\\\
    11 & University of Maryland - College Park &  36& Ohio State \\
    12 & Northeastern University\red{$^*$} &37& University of Virginia  \\
    13 & Purdue University &38& Yale University \\
    14 & University of Wisconsin - Madison\red{$^*$} &39 & Univ. of California - Santa Cruz \\
    15 & University of Texas at Austin &40& Brown University \\
    16 & University of Pennsylvania &41 & Harvard University \\
    17 & Columbia University\red{$^*$} &42 & Boston University  \\
    18 & Princeton University\red{$^*$}  & 43& Rice University\\
    19 & New York University  & 44&  University at Buffalo\red{$^*$}\\
    20 &  University of Massachusetts-Amherst\red{$^*$} &45& University of Colorado-Boulder \\
    21 & Univ. of California - Los Angeles &46& University of Illinois at Chicago  \\
    22 & University of Southern California &47& University of Minnesota \\
    23 & University of Chicago &48& University of North Carolina\red{$^*$} \\
    24 & Stony Brook University\red{$^*$} &49& Arizona State University\red{$^*$} \\
    25 &  Univ. of California - Irvine&50& Univ. Of California - Davis \\
    \bottomrule
  \end{tabular}
\end{table}
% Virginia Tech\red{$^*$}  \\

>>>>

Note that most places only have deadlines for the applicant, but are very flexible with the letter writers.  In many cases your LOR writers \emph{are not given any deadline}. \autoref{fig:lor-invitations} lists several LOR invitation emails I received from various universities in the past few years.

@@

    \textbf{Sample 1:}
    .. is applying for admission into the Computer Science (Ph.D.) program in the Graduate School at the University of Massachusetts Amherst and has listed you as a reference.

@@

    \textbf{Sample 2:}
    Dear ..

@@

    \textbf{Sample 3:}
    .. is applying to the University of Nebraska-Lincoln and has named you as a recommender. \textbf{This applicant has waived the right to view your recommendation.}

@@

    \textbf{Sample 2:}
    Dear Recommender,

@@

The reason is similar to LoRs (\autoref{sec:ievaluate}). If we know your school has a good reputation, then we can trust the education and grades from your school.  If we do not know your school, then we are uncertain about its quality and the grades you received.

Many international students mistakenly believe that their school is well-known, but in fact, it is not, e.g., many Vietnamese students believe that ``Bach Khoa'', which is one of the best universityies in the country, is well-known, but it is not (and their confusing acronyms HUST and HCMUT only make it worse).  So if you think your school is well-known, mention it (and back up with some evidence like rankings or awards) in your SOP.  You can ask your LOR writers to talk about your school as well (\autoref{sec:help-your-LOR-writers}).
Of course, if you're interested in working with Vietnamese, consider  \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{CS programs in the US that have Vietnamese professors}. %It might also be helpful to have your CS dept to put itself on CSRankings so that others know about the dept, its faculty, and research.

@@

\section{Locally}

Start looking for research opportunities at your institution.
If you did well or liked a class, you can check with the professor of that class for research opportunities.
You can also go to the department directory and then professors' websites and see if they are looking for undergraduate researchers.
Even if they say they are looking for graduate students, you can still contact them and ask.

Many universities have programs to encourage undergraduate research. For example, Mason has the OSCAR program, UNL has UCARE, and NSF has REUs (for US citizens and permanent residents).

@@

\section{Open Source Contributions}

Places such as GitHub offer many research opportunities, even though they might not appear so (e.g., they do not explicitly mention research).
In many cases, professors or research labs put their projects on GitHub. For example, my research group, \href{https://github.com/dynaroars/}{Dynaroars}, has many open-source projects that undergrads can contribute to.

@@

\section{Virtual Research Opportunities} 
This is less common but several places offer virtual internships and research programs aimed at providing hands-on research experience. These programs often involve working remotely under the guidance of experienced mentors and collaborating with a team of fellow researchers. For instance, \href{https://docs.google.com/forms/d/1btIwt4HwjyKMOUk-EMy3rbkfWzFxv2lNrMm_zkd0pA4/viewform?edit_requested=true}{UIUC+ Summer Undergraduate Research in Software Engineering}  offers an unpaid remote internship for software engineering students all over the world to collaborate with mentors from University of Illinois at Urbana Champaign.

@@

\begin{table}
  \centering
  \small
  \caption{The top 50 CS programs in the US from \href{https://www.csrankings.org}{CSRankings.org}, a ranking system  based on faculty publications at top CS conferences. (\href{https://csrankings.org}{CSRankings}, Jan. 2024). \red{$^*$} indicates that the university has \href{https://github.com/dynaroars/dynaroars.github.io/wiki/Viet-CS-Profs-US}{Vietnamese prof.} that can advise CS PhD students.}\label{tab:ranking}
  \begin{tabular}{rl|rl}
    \toprule
    1 & Carnegie Mellon & 26 & Duke University \\
    2 & Univ. of Illinois at Urbana-Champaign\red{$^*$}  & 27 & Univ. of California - Santa Barbara \\
    3 & Univ. of California-San Diego & 28 & Rutgers University\red{$^*$} \\
    4 & Georgia Institute of Technology & 29 & Univ. of California - Riverside\\
    5 & MIT                            & 30 & Pennsylvania State University  \\
    6 & Univ. of California - Berkeley& 31 & Northwestern University\\
    7 & University of Michigan - Ann Arbor\red{$^*$}   & 32& George Mason University\red{$^*$}\\
    8 & University of Washington      &33 &  Texas A\&M University\red{$^*$} \\
    9 &  Stanford University  &34&  University of Utah \\
    10 & Cornell University  & 35 &  North Carolina State University\\\
    11 & University of Maryland - College Park &  36& Ohio State \\
    12 & Northeastern University\red{$^*$} &37& University of Virginia  \\
    13 & Purdue University &38& Yale University \\
    14 & University of Wisconsin - Madison\red{$^*$} &39 & Univ. of California - Santa Cruz \\
    15 & University of Texas at Austin &40& Brown University \\
    16 & University of Pennsylvania &41 & Harvard University \\
    17 & Columbia University\red{$^*$} &42 & Boston University  \\
    18 & Princeton University\red{$^*$}  & 43& Rice University\\
    19 & New York University  & 44&  University at Buffalo\red{$^*$}\\
    20 &  University of Massachusetts-Amherst\red{$^*$} &45& University of Colorado-Boulder \\
    21 & Univ. of California - Los Angeles &46& University of Illinois at Chicago  \\
    22 & University of Southern California &47& University of Minnesota \\
    23 & University of Chicago &48& University of North Carolina\red{$^*$} \\
    24 & Stony Brook University\red{$^*$} &49& Arizona State University\red{$^*$} \\
    25 &  Univ. of California - Irvine&50& Univ. Of California - Davis \\
    \bottomrule
  \end{tabular}
\end{table}
% Virginia Tech\red{$^*$}  \\
